---
title: "Parent Paper 1: Data Split"
output: html_notebook
---

TEST FOR GITHUB REPO

DS 340W: Fall 2026
Maeric Barrows & Ayla Orona

This is documentation for processes performed for Parent Paper 1.

# 0. Load Data and Packages

```{r}
remove(list=ls())

library(dplyr)
library(tidyverse)

library(rsample) # Data split
library(pROC) # ROC curves
library(smotefamily) # SMOTE
library(caret) # Regression
library(ggplot2)
library(corrplot)

data <- read.csv("parentpaper1_data.csv")
```

# 1. Training/Testing/Validation Split

```{r Data Cleaning}
# Convert binary variables (1= No, 2 = Yes) to 0/1
binary_vars <- c("SMOKING", "YELLOW_FINGERS", "ANXIETY", "PEER_PRESSURE", "CHRONIC.DISEASE", "FATIGUE", "ALLERGY", "WHEEZING", "ALCOHOL.CONSUMING", "COUGHING","SHORTNESS.OF.BREATH", "SWALLOWING.DIFFICULTY", "CHEST.PAIN")

data[binary_vars] <- lapply(data[binary_vars], function(x) ifelse(x == 2, 1, 0))

# Convert gender to binary(M = 1, F = 0)
data$GENDER <- ifelse(data$GENDER == "M", 1, 0)

# Convert Lunge cancer into binary (YES = 1,  N0 = 0)

data$LUNG_CANCER <- ifelse(data$LUNG_CANCER == "YES", 1, 0)
```


```{r Data Split}
split <- initial_validation_split(data, prop=c(0.7, 0.1))

train_set <- training(split)
test_set <- testing(split)
validation_set <- validation(split)
```

```{r}
# For submission
write.csv(train_set, "training.csv")
write.csv(test_set, "testing.csv")
write.csv(validation_set, "validation.csv")
```


## Aylas work
## Load Libraries


```{r}
# Load the necessary libraries
library(pROC)
library(smotefamily)
library(caret) # for regression
library(ggplot2)
library(corrplot)
library(tidyverse)
library(dplyr)
library(randomForest)
#library(FSelector)  # for Gain Ratio but not working so computed it manually instead
library(themis)
library(kableExtra)
```

## Load the data

```{r}
# Load the dataset
lung_cancer <- read.csv("survey lung cancer.csv")

head(lung_cancer)
```

## Data Cleaning

```{r}
# Convert binary variables (1= No, 2 = Yes) to 0/1
binary_vars <- c("SMOKING", "YELLOW_FINGERS", "ANXIETY", "PEER_PRESSURE", "CHRONIC.DISEASE", "FATIGUE", "ALLERGY", "WHEEZING", "ALCOHOL.CONSUMING", "COUGHING","SHORTNESS.OF.BREATH", "SWALLOWING.DIFFICULTY", "CHEST.PAIN")

lung_cancer[binary_vars] <- lapply(lung_cancer[binary_vars], function(x) ifelse(x == 2, 1, 0))

# Convert gender to binary(M = 1, F = 0)
lung_cancer$GENDER <- ifelse(lung_cancer$GENDER == "M", 1, 0)

# Convert Lunge cancer into binary (YES = 1,  N0 = 0)

lung_cancer$LUNG_CANCER <- ifelse(lung_cancer$LUNG_CANCER == "YES", 1, 0)

# Check for any missing values
colSums(is.na(lung_cancer))


# check the structure of the data
str(lung_cancer)
```

## Exploratory Data Analysis

```{r}

# Summarry statistics
summary(lung_cancer)

# Proportion of lung cancer
prop.table(table(lung_cancer$LUNG_CANCER)) # (Class imbalance)
```

## 3.2 Data Preprocessing: SMOTE Class Balancing (50-50)
```{r}
# ============================================================
# 3.2 Data Preprocessing: SMOTE Class Balancing (50-50)
# ============================================================


# Convert categorical predictors to factors
# AGE remains numeric
smote_data <- lung_cancer %>%
  mutate(
    GENDER = as.factor(GENDER),
    SMOKING = as.factor(SMOKING),
    YELLOW_FINGERS = as.factor(YELLOW_FINGERS),
    ANXIETY = as.factor(ANXIETY),
    PEER_PRESSURE = as.factor(PEER_PRESSURE),
    CHRONIC.DISEASE = as.factor(CHRONIC.DISEASE),
    FATIGUE = as.factor(FATIGUE),
    ALLERGY = as.factor(ALLERGY),
    WHEEZING = as.factor(WHEEZING),
    ALCOHOL.CONSUMING = as.factor(ALCOHOL.CONSUMING),
    COUGHING = as.factor(COUGHING),
    SHORTNESS.OF.BREATH = as.factor(SHORTNESS.OF.BREATH),
    SWALLOWING.DIFFICULTY = as.factor(SWALLOWING.DIFFICULTY),
    CHEST.PAIN = as.factor(CHEST.PAIN),
    LUNG_CANCER = as.factor(LUNG_CANCER)
  )

# Check original class distribution
table(smote_data$LUNG_CANCER)


# ============================================================
# Apply SMOTENC
# ============================================================

set.seed(123)

lung_cancer_balanced <- smotenc(
  smote_data,
  var = "LUNG_CANCER",
  k = 5,
  over_ratio = 1
)


# ============================================================
# Check balanced dataset
# ============================================================

table(lung_cancer_balanced$LUNG_CANCER)

prop.table(table(lung_cancer_balanced$LUNG_CANCER))

dim(lung_cancer_balanced)

str(lung_cancer_balanced)
```

## 3.3 Features Analysis: Age Distribution on Balanced Data

```{r}
# Create age groups matching 5-year bins
# Widen lower bound to avoid NA bin from any outlier ages
lung_cancer_balanced <- lung_cancer_balanced %>%
  mutate(AGE_GROUP = cut(AGE,
                          breaks = c(-Inf, 39, 44, 49, 54, 59, 64, 69, 74, 79, 84, Inf),
                          labels = c("35 -- 39", "40 -- 44", "45 -- 49", "50 -- 54",
                                     "55 -- 59", "60 -- 64", "65 -- 69", "70 -- 74",
                                     "75 -- 79", "80 -- 84", "85+"),
                          right = TRUE))

# Confirm no NAs remain
sum(is.na(lung_cancer_balanced$AGE_GROUP))

# Calculate percentage as share of the WHOLE balanced dataset (not within-group)
age_group_pct <- lung_cancer_balanced %>%
  count(LUNG_CANCER, AGE_GROUP, .drop = FALSE) %>%
  mutate(pct = n / sum(n)) %>%   # denominator = total rows in balanced data
  mutate(LUNG_CANCER = factor(LUNG_CANCER, levels = c(0, 1), labels = c("No", "Yes")))

# Plot: grouped bar chart with percentages (matches paper's Figure 1)
ggplot(age_group_pct, aes(x = AGE_GROUP, y = pct, fill = LUNG_CANCER)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01),
                      limits = c(0, 0.18)) +
  scale_fill_manual(values = c("No" = "#4472C4", "Yes" = "#ED7D31")) +
  labs(title = "Lung Cancer", 
       x = "Age group", y = "Lung cancer", fill = "LUNG_CANCER") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_text(size = 10),
        legend.position = "top",
        legend.justification = "right")
```


## 3.3 Features Analysis: Feature Ranking (Table 1)

```{r}
# ============================================================
# Random Forest Feature Importance (Gini-based)
# ============================================================

ranking_data <- lung_cancer_balanced

set.seed(123)

rf_model <- randomForest(
  LUNG_CANCER ~ .,
  data = ranking_data,
  importance = TRUE,
  ntree = 500
)

rf_importance <- importance(rf_model, type = 2) %>%
  as.data.frame() %>%
  rownames_to_column("Feature") %>%
  rename(RF_Ranking = MeanDecreaseGini) %>%
  arrange(desc(RF_Ranking))

rf_importance
```

```{r}
# ============================================================
# Manual Gain Ratio calculation (no Java/FSelector dependency needed)
# Formula: GR(fi) = [H(c) - H(c|fi)] / H(fi)
# ============================================================

# Entropy function: H(X) = -sum(p * log2(p))
entropy <- function(x) {
  probs <- table(x) / length(x)
  probs <- probs[probs > 0]  # avoid log(0)
  -sum(probs * log2(probs))
}

# Conditional entropy function: H(c|f) = sum over f's values of P(f=v) * H(c | f=v)
conditional_entropy <- function(class_var, feature_var) {
  levels_f <- unique(feature_var)
  total_n <- length(feature_var)
  
  cond_h <- 0
  for (v in levels_f) {
    subset_class <- class_var[feature_var == v]
    weight <- length(subset_class) / total_n
    cond_h <- cond_h + weight * entropy(subset_class)
  }
  return(cond_h)
}# Gain Ratio function for one feature
gain_ratio_single <- function(class_var, feature_var) {
  h_c <- entropy(class_var)
  h_c_given_f <- conditional_entropy(class_var, feature_var)
  h_f <- entropy(feature_var)
  
  info_gain <- h_c - h_c_given_f
  
  # Avoid divide-by-zero if feature has only 1 unique value
  if (h_f == 0) return(0)
  
  gr <- info_gain / h_f
  return(gr)
}
```


```{r}
# ============================================================
# Apply Gain Ratio to all features
# ============================================================

feature_names <- setdiff(names(ranking_data), "LUNG_CANCER")
class_var <- ranking_data$LUNG_CANCER

gr_scores <- sapply(feature_names, function(f) {
  gain_ratio_single(class_var, ranking_data[[f]])
})

gain_ratio_result <- data.frame(
  Feature = names(gr_scores),
  GR_Ranking = round(as.numeric(gr_scores), 4)
) %>%
  arrange(desc(GR_Ranking))

gain_ratio_result
```

```{r}
# ============================================================
# Combine into Table 1 format (Random Forest | Gain Ratio side by side)
# ============================================================

table1 <- data.frame(
  RF_Feature = rf_importance$Feature,
  RF_Ranking = round(rf_importance$RF_Ranking, 4),
  GR_Feature = gain_ratio_result$Feature,
  GR_Ranking = gain_ratio_result$GR_Ranking
)

table1
```


# TABLE 2 WORK
```{r}
# ============================================================
# Table 2: Feature values distribution by Lung Cancer class
# ============================================================

# List of binary features to summarize (excludes AGE - it's continuous)
binary_features <- c("GENDER", "SMOKING", "YELLOW_FINGERS", "ANXIETY", 
                      "PEER_PRESSURE", "CHRONIC.DISEASE", "FATIGUE", "ALLERGY", 
                      "WHEEZING", "ALCOHOL.CONSUMING", "COUGHING", 
                      "SHORTNESS.OF.BREATH", "SWALLOWING.DIFFICULTY", "CHEST.PAIN")

# Force all binary feature columns to numeric before rounding
# (as.character first avoids the classic factor -> wrong-number bug)
lung_cancer_balanced_rounded <- lung_cancer_balanced %>%
  mutate(across(all_of(binary_features), 
                ~ round(as.numeric(as.character(.)))))

# Confirm it worked
str(lung_cancer_balanced_rounded %>% select(all_of(binary_features)))
```

```{r}
# Function: for one feature, compute % of whole dataset in each 
# Feature-value x LUNG_CANCER combination
feature_table2 <- function(feature_name, data) {
  data %>%
    count(!!sym(feature_name), LUNG_CANCER, .drop = FALSE) %>%
    mutate(pct = n / sum(n) * 100) %>%
    mutate(LUNG_CANCER = factor(LUNG_CANCER, levels = c(0, 1), labels = c("No", "Yes"))) %>%
    select(!!sym(feature_name), LUNG_CANCER, pct) %>%
    pivot_wider(names_from = LUNG_CANCER, values_from = pct) %>%
    rename(Value = !!sym(feature_name)) %>%
    mutate(Feature = feature_name, .before = Value)
}

# Apply to every binary feature and combine into one long table
table2_long <- map_dfr(binary_features, ~ feature_table2(.x, lung_cancer_balanced_rounded))

table2_long
```

```{r}
# ============================================================
# Format Value labels (0/1 -> No/Yes, or Female/Male for gender)
# ============================================================

table2_formatted <- table2_long %>%
  mutate(Value = case_when(
    Feature == "GENDER" & Value == 0 ~ "Female",
    Feature == "GENDER" & Value == 1 ~ "Male",
    Value == 0 ~ "No",
    Value == 1 ~ "Yes",
    TRUE ~ as.character(Value)
  )) %>%
  mutate(No = round(No, 2), Yes = round(Yes, 2))

table2_formatted
```

```{r}
# ============================================================
# Display as a clean kable table (long format, one feature block at a time)
# ============================================================

kable(table2_formatted, 
      col.names = c("Feature", "Value", "No", "Yes"),
      caption = "Table 2. The distribution of participants in terms of feature values and class label in the balanced data.",
      align = "c") %>%
  kable_styling(full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r}
# ============================================================
# Optional: side-by-side two-column layout like the paper 
# (splits features into two halves, binds them next to each other)
# ============================================================

features_left <- binary_features[1:7]
features_right <- binary_features[8:14]

table2_left <- table2_formatted %>% filter(Feature %in% features_left)
table2_right <- table2_formatted %>% filter(Feature %in% features_right)

# Bind side by side (row counts match since each feature has 2 rows: No/Yes value)
table2_side_by_side <- bind_cols(
  table2_left %>% rename(Feature_L = Feature, Value_L = Value, No_L = No, Yes_L = Yes),
  table2_right %>% rename(Feature_R = Feature, Value_R = Value, No_R = No, Yes_R = Yes)
)

kable(table2_side_by_side,
      col.names = c("Feature", "Value", "No", "Yes", "Feature", "Value", "No", "Yes"),
      caption = "Table 2. The distribution of participants in terms of feature values and class label in the balanced data.",
      align = "c") %>%
  kable_styling(full_width = FALSE) %>%
  add_header_above(c(" " = 2, "Lung Cancer" = 2, " " = 2, "Lung Cancer" = 2))
```


